Mergeable Types for High Availablity

### 2017-02-21 paper/model.tex

  \begin{ocaml}
  module type DALI = sig
    type 'a 'b t
    val return : 'b -> 'a 'b t
    val bind : 'a 'b t -> ('b -> 'a 'c t) -> 'a 'c t
    val get_current_version: unit -> 'a 'a t
    val with_init_version_do: 'a -> 'a 'b t -> 'b
    val fork_version : 'a 'b t -> 'a unit t
    val sync_next_version: unit -> ?v:'a -> 'a 'a t
  end
  \end{ocaml}

@@

@@

\name requires a three-way \C{merge} function to merge the concurrent
versions of a drawing canvas. The three arguments include two
concurrent versions (\C{v1} and \C{v2}), and their least common
ancestor (\C{lca}) - the version from which the two concurrent
versions evolved independently. The merge function can make use of the
pixel values of the common ancestor to merge the pixel values on both
the canvases. For instance, if \C{v1} colors the pixel at $(0,0)$ red,
and \C{v2} colors it green, and the color of the pixel in \C{lca} is
red, then it means that only \C{v2} modified the color. Hence the
pixel is colored green in the merged canvas. On the other hand, if the
pixel is colored white in the original canvas, then it means that both
\C{v1} and \C{v2} have modified the color. In such case, an appopriate
color mixing algorithm can be used to determine the color of pixel.
For instance, the pixel can be colored yellow - an additive
combination of red and green.

The \name programming model lets programmers define and compose
concurrent computations around versioned data structures.
Fig.~\ref{fig:dali-monad} shows the signature of the \name module that
implements the programming model along the lines of the well-known
\C{State} monad~\cite{wadler-monad}. The monad encapsulates a
versioned functional state (\C{'a}) and the type \C{'a 'b t}
represents a monadic computation that returns a \C{'b} result.
Functions \C{return} and \C{bind} have their usual monadic
interpretation. \C{get\_current\_version} is like the \C{State}
monad's \C{get}; it returns the versioned state encapsulated by the
monad. \C{with\_init\_version\_do} runs a monadic computation against
a given initial version and returns the result. \C{fork\_version}
returns a computation that forks a new concurrent version from the
current version, and runs the given monadic computation asynchronously
against the forked version.  The semantics of \C{sync\_next\_version}
(simply called \C{sync}) is somewhat involved: the function accepts a
\emph{proposal} for the next version of the state, where \C{next} is
defined w.r.t the current version, and returns (via a monad) the
actual next version, which becomes the current version for rest of the
computation.  The next version is created by merging the proposal with
a subset of concurrent versions that became available since the last
merge or fork. Thus, \C{sync} effectively lets a computation sync with
the concurrent computations and obtain their latest updates.

\caption{\drawsome: a sample collaborative }
\label{fig:dali-canvas}
\end{figure}

>>>>

  \begin{ocaml}
  module type DALI = sig
    type 'a 'b t
    val return : 'b -> 'a 'b t
    val bind : 'a 'b t -> ('b -> 'a 'c t) -> 'a 'c t
    val get_current_version: unit -> 'a 'a t
    val with_init_version_do: 'a -> 'a 'b t -> 'b
    val fork_version : 'a 'b t -> 'a unit t
    val sync ()_next_version: unit -> ?v:'a -> 'a 'a t
  end
  \end{ocaml}

@@

\begin{figure}
\centering
\includegraphics[scale=0.5]{Figures/canvas-merging}

\caption{Merging concurrent versions (\C{v1} and \C{v2} ) of a drawing
canvas. \C{lca} is their common ancestor.}
\label{fig:canvas-merging}
\end{figure}

@@

\name requires a three-way \C{merge} function to merge the concurrent
versions of a drawing canvas. The three arguments include two
concurrent versions (\C{v1} and \C{v2}), and their least common
ancestor (\C{lca}) - the version from which the two concurrent
versions evolved independently. The merge function can make use of the
pixel values of the common ancestor to merge the pixel values on both
the canvases. For instance, if the color of a pixel in \C{v1} is white,
and in \C{v2} it is green, and its color in \C{lca} is
white, then it means that only \C{v2} modified the color. Hence the
pixel is colored green in the merged canvas. On the other hand, if the
pixel is red in \C{v1}, then it means that both
\C{v1} and \C{v2} have modified the color. In such case, an appopriate
color-mixing algorithm can be used to determine the color of pixel.
For instance, the pixel can be colored yellow - an additive
combination of red and green. The logic is illustrated in
Fig.~\ref{fig:canvas-merging}.

The \name programming model lets programmers define and compose
concurrent computations around versioned data structures.
Fig.~\ref{fig:dali-monad} shows the signature of the \name module that
implements the programming model along the lines of the well-known
\C{State} monad~\cite{wadler-monad}. The monad encapsulates a
versioned functional state (\C{'a}) and the type \C{'a 'b t}
represents a monadic computation that returns a \C{'b} result.
Functions \C{return} and \C{bind} have their usual monadic
interpretation. \C{get\_current\_version} is like the \C{State}
monad's \C{get}; it returns the versioned state encapsulated by the
monad. \C{with\_init\_version\_do} runs a monadic computation against
a given initial version and returns the result. \C{fork\_version}
returns a computation that forks a new concurrent version from the
current version, and runs the given monadic computation async ()hronously
against the forked version.  The semantics of \C{sync ()\_next\_version}
(simply called \C{sync ()}) is somewhat involved: the function accepts a
\emph{proposal} for the next version of the state, where \C{next} is
defined w.r.t the current version, and returns (via a monad) the
actual next version, which becomes the current version for rest of the
computation.  The next version is created by merging the proposal with
a subset of concurrent versions that became available since the last
merge or fork. Thus, \C{sync ()} effectively lets a computation sync
() with a subset of concurrent computations and obtain their latest
updates.

\begin{figure}
\centering
\begin{tabular}{l||l||l}
\begin{ocaml}
let alice_f : C.t unit t = 
  get () >>= fun c0 -> 
  fork bob_f >>= fun () ->
  let c0' = C.draw_line c0 
    {x=10;y=20}
    {x=40;y=20} in
  sync () ~v:c0' >>= 
  fun c1 ->
  let c1' = C.draw_line c1 
    {x=10;y=40} 
    {x=40;y=40} in
  sync () ~v:c1' >>= 
  fun c2 -> return ()
\end{ocaml}
&
\begin{ocaml}
let bob_f : C.t unit t = 
  get () >>= fun c0 -> 
  fork cheryl_f >>= 
  fun () ->
  let c0' = C.draw_line c0 
    {x=10;y=20} 
    {x=10;y=40} in
  sync () ~v:c0' >>= 
  fun c1 -> sync () >>= 
  fun c2 -> return ()
\end{ocaml}
&
\begin{ocaml}
let cheryl_f : C.t unit t = 
  get () >>= fun c0 -> 
  let c0' = C.draw_line c0 
    {x=40;y=20} 
    {x=40;y=20} in
  sync () ~v:c0' >>= 
  fun c1 -> sync () >>= 
  fun c2 -> return ()
\end{ocaml}
\\
\end{tabular}
\caption{\drawsome: A collaborative drawing session between Alice,
Bob, and Cheryl}
\label{fig:canvas-sessions-code}
\end{figure}

Fig.~\ref{fig:canvas-sessions} shows a collaborative editing session
between Alice, Bob and Cheryl.

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/canvas-sessions}

\caption{\drawsome: Collaborative drawing session visualized}
\label{fig:canvas-sessions}
\end{figure}
